---
title: "Reproducing Results for Paper: Fast Variational Bayes for Large Spatial Data"
author: "Jiafang Song, Abhirup Datta"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

# Overview

This document describes how to reproduce all simulation, real-world, and supplementary results in the paper **Fast Variational Bayes for Large Spatial Data**.

## Get only the `paper/` folder
If only need the manuscript files in `paper/` and do not want to check out the full repository, use Git sparse checkout.
```bash
git clone --filter=blob:none --no-checkout https://github.com/jfsong100/spVarBayes.git
cd spVarBayes
git sparse-checkout init --cone
git sparse-checkout set paper
git checkout main
```

The workflow has the following components:

1. **Environment setup**
2. **Simulation studies**  
   (Figures 1–3, Table 3)
3. **Real-world data analysis**  
   (Figures 4–7, Table 4)
4. **Supplementary experiments**  
   (Figures G.1–G.2, J.1–J.2, Figures H.1–H.6, I.1–I.2)

> **Note:** For Large‐scale settings (`n_index = 4, 5`) and methods **DKLGP**, **VNNGP**, and **spVB-MFA-LR** require substantial memory.

# Environment Setup

## R / spVarBayes

Install the `spVarBayes` package from GitHub:

```{r install-spVarBayes, eval=FALSE}
# Install devtools if needed
install.packages("devtools")

# Install spVarBayes
devtools::install_github("jfsong100/spVarBayes")

# Install spNNGP
install.packages("spNNGP")
```

This package provides the following methods:

- `spVB-MFA`
- `spVB-MFA-LR`
- `spVB-NNGP`
- `spVB-NNGP-joint`

Make sure the R version and system compilers are configured so that the package (and its C++/Rcpp components) install successfully.

## Python / VNNGP Environment

We recommend a dedicated conda environment for **VNNGP**.

```bash
# Create and activate environment
conda create -n env_vnngp python=3.10
conda activate env_vnngp

# Core dependencies
conda install pytorch==1.13.1
conda install gpytorch==1.10
conda install conda-forge::nb_conda
conda install notebook==6.5.6
conda install jupyter
conda install h5py
conda install pandas
conda install matplotlib
```

Use this environment when running `VNNGP.py` (both simulations and real-data analysis).

## Python / DKLGP Environment

Create a separate conda environment for **DKLGP**:

```bash
# Create and activate environment
conda create -n env_dklgp python=3.10
conda activate env_dklgp
```

Clone and install the DKLGP package:

```bash
# In a folder where you want the DKL-GP source
git clone https://github.com/katzfuss-group/DKL-GP.git

# Direct to that folder (change this)
cd ~/DKL-GP

# Install DKLGP and its Python dependencies
source INSTALL
```

Install Jupyter-related dependencies:

```bash
conda install conda-forge::nb_conda
conda install notebook==6.5.6
conda install jupyter
conda install h5py
conda install matplotlib
```

Use this environment when running `DKLGP.py` and `DKLGP_default.py`.

# Simulation Studies

The simulation studies in the paper consider sample sizes $n = 1000,5000,10000,50000,100000$, with the corresponding n_index values $1,2,3,4,5$. For a quick code check, we recommend running `n_index=3`, which reproduces the results for $n=10000$. To fully reproduce the paper’s simulation results, run all n_index values and set the seed from 1 to 100.

## Data Generation

**Script:** `data_generation.R`

- **Inputs:**
  - `n_index` (controls sample size)
  - `t` (seed)
- **Usage (example from shell):**

```bash
## Example: generate data for n_index = 3 t = 1
Rscript data_generation.R n_index=3 t=1
```

```bash
## Example: generate data for n_index = 1, 2, 3 and t = 1, 2
for n_index in 1 2 3; do
  for t in 1 2; do
    Rscript data_generation.R n_index=${n_index} t=${t}
  done
done
```

- **Output:**
  - Creates a folder: `data_sim/`
  - For each combination of `n_index` and `t`, saves simulated data.
  - For `n_index = 1, 2, 3`:
    - Also saves the empirical covariance matrix (large files).
  - For `n_index = 4, 5`:
    - Empirical covariance is **not** saved due to memory limits.
    - Only outcomes, covariates, and coordinates are saved.

Details for each scenario are described in Section 4 of the paper.

## R-based Methods: spVarBayes + spNNGP

**Script:** `spVB_spNNGP.R`

- **Inputs:**
  - `n_index`
  - `t` (seed)
- **Usage (example):**

```bash
## Example: run analysis for n_index = 3 t = 1
Rscript spVB_spNNGP.R n_index=3 t=1
```

```bash
# e.g. for multiple runs
for n_index in 1 2 3; do
  for t in 1 2; do
    Rscript spVB_spNNGP.R n_index=${n_index} t=${t}
  done
done
```

- **What it does:**
  - Reads data from `data_sim/`.
  - Runs:
    - `spVB-MFA`
    - `spVB-MFA-LR`
    - `spVB-NNGP`
    - `spVB-NNGP-joint`
    - `spNNGP`
  - Creates a folder: `R_results/`.

- **Saved outputs in `R_results/` include:**
  - **Fixed effects:**
    - `output_beta_vector`: point estimates of $\beta$
    - `CI_beta`: confidence intervals for $\beta$
  - **Hyperparameters:**
    - `output_theta_vector`: point estimates for $\theta=\{\sigma^2,\tau^2,\phi\}$
  - **Random effects summaries:**
    - `output_list`:
      - Empirical mean and variance
      - Variational approximated mean and variance
      - spNNGP approximated mean and variance
  - **Coverage and interval metrics for random effects:**
    - `CI_w`: confidence intervals, coverage, CRPS, interval scores
  - **KL and timing (per method):**
    - `output_vector`: summarizes KL divergence and running time

## DKLGP Simulations

**Script:** `DKLGP.py` `DKLGP_default.py`

- **Note:** the only difference between `DKLGP.py` and `DKLGP_default.py` is convergence options inside the file
  - **DKLGP-default:** set `converged = FALSE`
  - **DKLGP:** set `converged = TRUE`

- **Environment:**

```bash
conda activate env_dklgp
```

- **Inputs:**
  - `n_index`
  - `seed`
  - A `setups.yaml` file in the **same folder** as `DKLGP.py` and `DKLGP_default.py`, containing default settings from the DKLGP paper.

- **Usage (example):**

```bash
python3 DKLGP.py <seed> <n_index>
# e.g.
python3 DKLGP.py 1 3
```

```bash
# e.g. for multiple runs
for n_index in 1 2 3; do
  for seed in 1 2; do
    python3 DKLGP.py ${seed} ${n_index}
    python3 DKLGP_default.py ${seed} ${n_index}
  done
done
```

```bash
conda deactivate
```

- **Output:**
  - Creates `DKLGP_results/` with:
    - $\theta$ point estimates
    - CRPS, interval scores, coverage for $w$
    - Running time (e.g., `KL_vec`)
    - `output_data` (DKLGP approximated mean/variance)

> **Memory note:** For `n_index = 4, 5`, DKLGP requires large memory.

## VNNGP Simulations

**Script:** `VNNGP.py`

- **Environment:**

```bash
conda activate env_vnngp
```

- **Inputs:**
  - `n_index`
  - `seed`

- **Usage (example):**

```bash
python3 VNNGP.py <seed> <n_index>
# e.g.
python3 VNNGP.py 1 3
```

```bash
# e.g. for multiple runs
for n_index in 1 2 3; do
  for seed in 1 2; do
    python3 VNNGP.py ${seed} ${n_index}
  done
done
```
```bash
conda deactivate
```

- **What it does:**
  - Reads data from `data_sim/`.
  - Runs VNNGP.

- **Output:**
  - Creates `VNNGP_results/` with:
    - θ point estimates
    - CRPS, interval scores, coverage for w
    - Running time (`KL_vec` or similar)
    - `output_data` (VNNGP approximated mean/variance)


> **Memory note:** `n_index = 4, 5` also require large memory for VNNGP.

## Simulation Summary and Figures

Once **all** `n_index` and seed combinations have been run for all methods (at least run n_index = 3):

1. **Combined simulation summary:**

   **Script:** `summary.R`
   This script uses `R_results/`, `DKLGP_results/`, and `VNNGP_results/` to create:

   ```bash
   Rscript summary.R
   ```

2. **This script creates:**
   - Figure 1, 2, 3
   - Figure H.1, H.2, H.3, H.4, H.5, H.6
   - Figure I.1, I.2  

Make sure the paths expected in `summary.R` match your folder structure.

# Real-World Data Analysis

## Data Processing

**Script:** `process_data.R`

- **Usage:**

```bash
Rscript process_data.R
```

- **Output:**
  - Creates a `data/` folder with:
    - Training data and Test data

## R-based Methods on Real Data

Run the following R scripts (each reads from `data/` and saves results accordingly):

```bash
Rscript spVB_MFA.R
Rscript spVB_MFA_LR.R
Rscript spVB_NNGP_ind.R
Rscript spVB_NNGP_joint.R
Rscript spNNGP.R
```

These fit:

- `spVB-MFA`
- `spVB-MFA-LR`
- `spVB-NNGP` (independent)
- `spVB-NNGP` (joint)
- `spNNGP`

> **Memory note:** `spVB-MFA-LR` require large memory for storing the whole covariance matrix

## DKLGP and VNNGP on Real Data

Run DKLGP (default) in the DKLGP environment:

```bash
conda activate env_dklgp
python3 DKLGP_default.py
conda deactivate
```

Run VNNGP in the VNNGP environment:

```bash
conda activate env_vnngp
python3 VNNGP.py
conda deactivate
```

Both scripts read from the processed `data/` folder and produce outputs compatible with the R-based methods.

## Real-Data Summary and Figures

1. **Combine DKLGP and VNNGP outputs:**

   **Script:** `summary_DKLGP_VNNGP.R`

   ```bash
   Rscript summary_DKLGP_VNNGP.R
   ```

   This stores combined results in an `output/` folder.

2. **Create real-data figures and Table 4:**

   **Script:** `summary.R` (real-data version)

   ```bash
   Rscript summary.R
   ```

   - Produces:
     - Figure 4,5,6,7  
     - Table 4  
   - Output is saved in the `fig/` folder.

# Supplementary Experiments

## Choice of Number of Nearest Neighbors (Figures G.1–G.2)

Folder: `G_choices_nn/`  
Script: `n_neighbor.R`

- **Usage:**

```bash
cd G_choices_nn
Rscript n_neighbor.R
```

- **Output:**
  - Figure G.1, G.2  

These illustrate different choices for the number of nearest neighbors for the NNGP variational family.

## Prediction Study (Figures J.1–J.2)

Folder: `J_predictions/`

1. **Data generation:**

   Script: `data_generation_prediction.R` (or the exact name in your folder)

   ```bash
   cd J_predictions
   for n_index in 1 2; do
      for t in 1 2; do
        Rscript data_generation_prediction.R n_index=${n_index} t=${t}
      done
   done
   ```

   Saves prediction-study data in `data_sim/` under `J_predictions/`.

2. **R-based methods:**

   Script: `spVB_spNNGP_pred.R`

   ```bash
   for n_index in 1 2; do
      for t in 1 2; do
        Rscript spVB_spNNGP_pred.R n_index=${n_index} t=${t}
      done
   done
   ```

   Runs all R-based methods for the prediction study.

3. **VNNGP and DKLGP:**

   ```bash
   # DKLGP (default version and converged version)
   conda activate env_dklgp
   for n_index in 1 2; do
      for seed in 1 2; do
        python3 DKLGP.py ${seed} ${n_index}
        python3 DKLGP_default.py ${seed} ${n_index}
      done
   done
   conda deactivate

   # VNNGP
   conda activate env_vnngp
   for n_index in 1 2; do
      for seed in 1 2; do
        python3 VNNGP.py ${seed} ${n_index}
      done
   done
   conda deactivate
   ```

4. **Summary and prediction figures:**

   Script: `summary.R` (prediction version)

   ```bash
   Rscript summary.R
   ```

   - Produces:
     - Figure J.1, J.2  

# Notes on Memory and Runtime

- **Large-scale simulations (`n_index = 4, 5`):**
  - Require large memory for R and Python.
  - **DKLGP**, **VNNGP**, and **spVB-MFA-LR** can be particularly demanding.
- On a cluster:
  - Request sufficient RAM in your job script (e.g. 150–300 GB, depending on `n_index` and method).
  - Running different methods and `n_index` combinations in parallel jobs can speed up the complete reproduction.
